---
title: "A Visual Walk Through NHANES"
author: "Farnaz"
date: "October 4, 2026"
output:
  html_document:
    theme: flatly
    highlight: kate
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "92%"
)
pacman::p_load(tidyverse, cowplot, ggpubr, DT, here)

nhanes_data <- read_csv(
  here("datasets", "cleaned_NHANES.csv"),
  show_col_types = FALSE
) %>%
  mutate(gender = factor(gender))

diet <- read_csv(
  here("datasets", "diet.csv"),
  show_col_types = FALSE
)
```

# Exercise 1 {.tabset}

## Two separate views

```{r exercise-1-separate, fig.width=8, fig.height=4.8}
age_gender <- ggplot(
  nhanes_data,
  aes(x = age, colour = gender, fill = gender)
) +
  geom_density(alpha = 0.25, linewidth = 0.9) +
  scale_colour_manual(values = c("#D55E00", "#0072B2")) +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  labs(
    title = "Age distributions are similar across genders",
    x = "Age (years)",
    y = "Density",
    colour = "Gender",
    fill = "Gender"
  ) +
  theme_classic()

ethnicity_mix <- ggplot(
  nhanes_data,
  aes(x = ethnicity_2, fill = gender)
) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  labs(
    title = "Gender counts across ethnicity groups",
    x = NULL,
    y = "Participants",
    fill = "Gender"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

age_gender
ethnicity_mix
```

The density plot emphasizes the shape of each age distribution. The grouped
bars answer a different question by showing the sample counts within each
ethnicity group.

## Shared-legend comparison

```{r exercise-1-combined, fig.width=10, fig.height=4.8}
ggpubr::ggarrange(
  age_gender,
  ethnicity_mix,
  ncol = 2,
  common.legend = TRUE,
  legend = "bottom"
)
```

For a single-page comparison, I prefer the shared legend because it gives the
long category labels more room. The separate display is easier when each plot
needs to stand on its own.

# Exercise 2

```{r exercise-2-table}
display_table <- nhanes_data %>%
  select(
    patient_id,
    wave_id,
    gender,
    age,
    ethnicity_2,
    education
  )

DT::datatable(
  display_table,
  rownames = FALSE,
  filter = "bottom",
  class = "stripe hover compact",
  options = list(
    pageLength = 8,
    autoWidth = TRUE,
    scrollX = TRUE
  ),
  caption = "Selected NHANES participant characteristics"
)
```

```{r exercise-2-figures, fig.width=8, fig.height=4.8}
age_plot <- ggplot(nhanes_data, aes(age, fill = gender)) +
  geom_density(alpha = 0.35) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
  labs(
    title = "Age profiles overlap across genders",
    x = "Age (years)",
    y = "Density",
    fill = "Gender"
  ) +
  theme_classic()

gender_plot <- nhanes_data %>%
  count(gender) %>%
  ggplot(aes(x = reorder(gender, n), y = n, fill = gender)) +
  geom_col(show.legend = FALSE, width = 0.68) +
  coord_flip() +
  scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
  labs(
    title = "The sample is balanced by gender",
    x = NULL,
    y = "Participants"
  ) +
  theme_light()

ethnicity_plot <- nhanes_data %>%
  count(ethnicity_2) %>%
  ggplot(aes(x = reorder(ethnicity_2, n), y = n)) +
  geom_col(fill = "#E67E22", width = 0.72) +
  coord_flip() +
  labs(
    title = "Detailed ethnicity composition",
    x = NULL,
    y = "Participants"
  ) +
  theme_classic()

ggsave(
  here("graphics", "farnaz_age_density.png"),
  age_plot,
  width = 7,
  height = 4.5
)
ggsave(
  here("graphics", "farnaz_gender_counts.png"),
  gender_plot,
  width = 7,
  height = 4.5
)
ggsave(
  here("graphics", "farnaz_ethnicity_detail.png"),
  ethnicity_plot,
  width = 7,
  height = 4.5
)

age_plot
gender_plot
ethnicity_plot
```

I kept the detailed ethnicity variable because it provides more information
about sample composition. The table supports exact lookups, while the ordered
bars and density curves make the overall patterns quicker to see.

# Exercise 3

```{r exercise-3, fig.width=8.5, fig.height=5}
diet_summary <- diet %>%
  group_by(Week) %>%
  summarise(
    mean_weight = mean(Weight),
    sd_weight = sd(Weight),
    .groups = "drop"
  )

ggplot(diet_summary, aes(Week, mean_weight)) +
  geom_ribbon(
    aes(
      ymin = mean_weight - sd_weight,
      ymax = mean_weight + sd_weight
    ),
    fill = "#F5B041",
    alpha = 0.30
  ) +
  geom_line(colour = "#C0392B", linewidth = 1.15) +
  geom_point(colour = "#C0392B", size = 2.2) +
  labs(
    title = "Average weight moves downward over 16 weeks",
    subtitle = "The shaded band is one standard deviation from the weekly mean",
    x = "Study week",
    y = "Mean weight"
  ) +
  theme_light()
```

The weekly summary removes the distracting participant legend. It keeps the
group trend as the main message and uses the ribbon to show that participant
weights still vary around the mean.

# Exercise 4

```{r exercise-4, fig.width=10, fig.height=5}
bp <- nhanes_data %>%
  mutate(
    readings_available = rowSums(
      !is.na(across(starts_with("systolic_bp_")))
    ),
    avg_sbp = rowMeans(
      across(starts_with("systolic_bp_")),
      na.rm = TRUE
    ),
    avg_sbp = if_else(
      readings_available == 0,
      NA_real_,
      avg_sbp
    ),
    sbp_group = case_when(
      avg_sbp < 120 ~ "Normal",
      avg_sbp < 130 ~ "Elevated",
      avg_sbp < 140 ~ "Stage 1",
      TRUE ~ "Stage 2"
    ),
    sbp_group = factor(
      sbp_group,
      levels = c("Normal", "Elevated", "Stage 1", "Stage 2")
    ),
    age_group = cut(
      age,
      breaks = seq(0, 90, by = 5),
      right = FALSE,
      include.lowest = TRUE
    )
  ) %>%
  filter(!is.na(age), !is.na(gender), !is.na(avg_sbp))

ggplot(bp, aes(age_group, fill = sbp_group)) +
  geom_bar(position = "dodge") +
  facet_grid(. ~ gender) +
  scale_fill_manual(
    values = c(
      "Normal" = "#56B4E9",
      "Elevated" = "#E69F00",
      "Stage 1" = "#D55E00",
      "Stage 2" = "#9B59B6"
    )
  ) +
  labs(
    title = "Higher SBP categories become more common with age",
    x = "Age group",
    y = "Participants",
    fill = "Average SBP"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The plot includes participants with age, gender, and at least one systolic blood
pressure reading. Facets reduce visual crowding, and the ordered colours help
the categories read from lower to higher blood pressure.

