---
title: "NHANES Wave Profiles"
subtitle: "A concise table report"
author: "Farnaz"
date: "October 4, 2026"
output:
  pdf_document:
    toc: true
    number_sections: false
    latex_engine: xelatex
bibliography: ../../citations/sources.bib
link-citations: true
header-includes:
  - \\usepackage{booktabs}
  - \\usepackage{longtable}
  - \\usepackage{array}
  - \\usepackage{float}
  - \\usepackage{xcolor}
  - \\definecolor{accent}{HTML}{C65D21}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  size = "\\scriptsize"
)
options(width = 68)
pacman::p_load(tidyverse, kableExtra, here)

nhanes_data <- read_csv(
  here("datasets", "cleaned_NHANES.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    wave_id = as.character(wave_id),
    readings_available = rowSums(
      !is.na(across(starts_with("systolic_bp_")))
    ),
    avg_sbp = rowMeans(
      across(starts_with("systolic_bp_")),
      na.rm = TRUE
    ),
    avg_sbp = if_else(
      readings_available == 0,
      NA_real_,
      avg_sbp
    )
  )

table_style <- function(table_object) {
  table_object %>%
    kable_styling(
      latex_options = c("striped", "hold_position"),
      full_width = FALSE,
      position = "center"
    ) %>%
    row_spec(0, bold = TRUE, color = "white", background = "#C65D21")
}
```

# Purpose

This report compares three NHANES waves using tables rather than figures. It
answers the requested questions about sample size, gender, ethnicity, and
systolic blood pressure. The cleaned data are derived from the National Health
and Nutrition Examination Survey [@cdc_nhanes].

# Sample size by wave

```{r sample-table}
sample_table <- nhanes_data %>%
  count(wave_id, name = "Participants") %>%
  arrange(wave_id)

kbl(
  sample_table,
  booktabs = TRUE,
  caption = "Number of participants in each survey wave",
  col.names = c("Wave", "Participants"),
  format.args = list(big.mark = ",")
) %>%
  table_style()
```

The 2015 wave has the largest sample. The number of participants is slightly
smaller in each later wave.

# Male participants

```{r male-table}
male_table <- nhanes_data %>%
  group_by(wave_id) %>%
  summarise(
    total = n(),
    male_n = sum(gender == "Male", na.rm = TRUE),
    male_percent = 100 * male_n / total,
    .groups = "drop"
  ) %>%
  mutate(
    male_percent = paste0(round(male_percent, 1), "%")
  ) %>%
  select(wave_id, male_n, male_percent)

kbl(
  male_table,
  booktabs = TRUE,
  caption = "Male count and percentage within each wave",
  col.names = c("Wave", "Male participants", "Percent male"),
  format.args = list(big.mark = ",")
) %>%
  table_style()
```

Male participants account for about half of each wave. The percentages are
very similar, so no wave has a noticeably different gender balance.

# Ethnicity composition

```{r ethnicity-table}
ethnicity_table <- nhanes_data %>%
  count(ethnicity_2, wave_id, name = "n") %>%
  group_by(wave_id) %>%
  mutate(
    percent = 100 * n / sum(n),
    result = paste0(
      format(n, big.mark = ","),
      " (",
      round(percent, 1),
      "%)"
    )
  ) %>%
  ungroup() %>%
  select(ethnicity_2, wave_id, result) %>%
  pivot_wider(names_from = wave_id, values_from = result)

kbl(
  ethnicity_table,
  booktabs = TRUE,
  caption = "Ethnicity counts with column percentages",
  col.names = c("Ethnicity", "2015", "2016", "2017")
) %>%
  table_style() %>%
  column_spec(1, width = "4.5cm")
```

The composition changes across waves, and the detailed ethnicity categories
make those shifts visible. Each cell shows the count followed by the percentage
within that wave.

# Average systolic blood pressure

```{r sbp-table}
sbp_table <- nhanes_data %>%
  filter(!is.na(avg_sbp)) %>%
  group_by(wave_id) %>%
  summarise(
    available_n = n(),
    mean = mean(avg_sbp),
    sd = sd(avg_sbp),
    median = median(avg_sbp),
    minimum = min(avg_sbp),
    maximum = max(avg_sbp),
    .groups = "drop"
  ) %>%
  mutate(
    `Mean (SD)` = paste0(round(mean, 1), " (", round(sd, 1), ")"),
    `Median [range]` = paste0(
      round(median, 1),
      " [",
      round(minimum, 1),
      ", ",
      round(maximum, 1),
      "]"
    )
  ) %>%
  select(wave_id, available_n, `Mean (SD)`, `Median [range]`)

kbl(
  sbp_table,
  booktabs = TRUE,
  caption = "Summary of participant-level average systolic blood pressure",
  col.names = c(
    "Wave",
    "Available n",
    "Mean (SD)",
    "Median [range]"
  ),
  format.args = list(big.mark = ",")
) %>%
  table_style()
```

The mean average systolic blood pressure is lowest in 2015 and highest in 2017.
The available sample is smaller than the full wave sample because participants
without any systolic blood pressure reading are excluded from this summary.

# Conclusion

The three waves have similar gender balance but differ in sample size,
ethnicity composition, and average systolic blood pressure. Presenting counts,
percentages, and summary statistics together gives both the scale and the
relative pattern of each comparison.

# References

