---
title: "NHAES Data Showcase"
author: "Farnaz"
date: "September 27, 2026"
output:
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  size = "\\scriptsize"
)
options(width = 70)
pacman::p_load(tidyverse, cowplot, ggpubr, here)

nhanes_data <- read_csv(
  here("datasets", "cleaned_NHANES.csv"),
  show_col_types = FALSE
)

diet <- read_csv(
  here("datasets", "diet.csv"),
  show_col_types = FALSE
)
```

# Exercise 1

```{r exercise-1, fig.width=9, fig.height=4.8}
nhanes_data <- nhanes_data %>%
  mutate(
    age_group = cut(age, breaks = seq(15, 90, 5), right = FALSE),
    gender = factor(gender)
  )

age_gender <- ggplot(
  nhanes_data,
  aes(x = age, colour = gender, fill = gender)
) +
  geom_density(alpha = 0.25) +
  labs(
    title = "Age distributions by gender",
    x = "Age (years)",
    y = "Density",
    colour = "Gender",
    fill = "Gender"
  ) +
  theme_classic()

ethnicity_mix <- ggplot(
  nhanes_data,
  aes(x = ethnicity_2, fill = gender)
) +
  geom_bar(position = "dodge") +
  labs(
    title = "Gender counts across detailed ethnicity",
    x = "Ethnicity",
    y = "Count",
    fill = "Gender"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

ggpubr::ggarrange(
  age_gender,
  ethnicity_mix,
  ncol = 2,
  common.legend = TRUE,
  legend = "bottom"
)
```

I prefer `ggarrange()` because the shared legend leaves more room for the two
plots. `plot_grid()` gives more control over relative widths, but separate
legends take space away from the longer ethnicity labels.

# Exercise 2

```{r exercise-2, fig.width=7.5, fig.height=5}
missing_summary <- nhanes_data %>%
  summarise(
    across(
      c(age, gender, ethnicity_1, ethnicity_2),
      ~ sum(is.na(.))
    )
  )
missing_summary

age_plot <- ggplot(nhanes_data, aes(age, fill = gender)) +
  geom_density(alpha = 0.35) +
  labs(
    title = "Age profiles overlap across genders",
    x = "Age (years)",
    y = "Density",
    fill = "Gender"
  ) +
  theme_classic()

gender_plot <- nhanes_data %>%
  count(gender) %>%
  ggplot(aes(x = reorder(gender, n), y = n, fill = gender)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Sample size by gender",
    x = "Gender",
    y = "Number of participants"
  ) +
  theme_light()

ethnicity_plot <- nhanes_data %>%
  count(ethnicity_2) %>%
  ggplot(aes(x = reorder(ethnicity_2, n), y = n)) +
  geom_col(fill = "#E67E22") +
  coord_flip() +
  labs(
    title = "Detailed ethnicity classification",
    x = "Ethnicity",
    y = "Count"
  ) +
  theme_classic()

age_plot
gender_plot
ethnicity_plot

ggsave(
  here("graphics", "farnaz_age_density.png"),
  age_plot,
  width = 7,
  height = 4.5
)
ggsave(
  here("graphics", "farnaz_gender_counts.png"),
  gender_plot,
  width = 7,
  height = 4.5
)
ggsave(
  here("graphics", "farnaz_ethnicity_detail.png"),
  ethnicity_plot,
  width = 7,
  height = 4.5
)
```

The density curves compare the shapes of the age distributions without letting
the larger gender group dominate only because it has more observations. The
ordered horizontal bars make the most and least frequent categories easy to
identify. I would keep `ethnicity_2` and remove `ethnicity_1`, because the more
detailed labels are more useful for describing the sample.

# Exercise 3

```{r exercise-3, fig.width=7.5, fig.height=4.8}
diet_summary <- diet %>%
  group_by(Week) %>%
  summarise(
    mean_weight = mean(Weight),
    sd_weight = sd(Weight),
    .groups = "drop"
  )

ggplot(diet_summary, aes(Week, mean_weight)) +
  geom_ribbon(
    aes(
      ymin = mean_weight - sd_weight,
      ymax = mean_weight + sd_weight
    ),
    fill = "#E69F00",
    alpha = 0.25
  ) +
  geom_line(colour = "#D55E00", linewidth = 1.1) +
  geom_point(colour = "#D55E00") +
  labs(
    title = "Mean weight declines over the 16-week study",
    subtitle = "Ribbon shows one standard deviation around the weekly mean",
    x = "Week",
    y = "Mean weight"
  ) +
  theme_light()
```

The original figure used one colour per participant, which produced a long
legend. I summarized by week so that the mean line communicates the typical
trajectory while the ribbon keeps information about participant variation.

# Exercise 4

```{r exercise-4, fig.width=9, fig.height=5}
bp <- nhanes_data %>%
  mutate(
    readings_available = rowSums(
      !is.na(across(starts_with("systolic_bp_")))
    ),
    avg_sbp = rowMeans(
      across(starts_with("systolic_bp_")),
      na.rm = TRUE
    ),
    avg_sbp = if_else(
      readings_available == 0,
      NA_real_,
      avg_sbp
    ),
    sbp_group = case_when(
      avg_sbp < 120 ~ "Normal",
      avg_sbp < 130 ~ "Elevated",
      avg_sbp < 140 ~ "Stage 1",
      TRUE ~ "Stage 2"
    ),
    age_group = cut(
      age,
      breaks = seq(0, 90, by = 5),
      right = FALSE,
      include.lowest = TRUE
    )
  ) %>%
  filter(!is.na(age), !is.na(gender), !is.na(avg_sbp))

ggplot(bp, aes(age_group, fill = sbp_group)) +
  geom_bar(position = "dodge") +
  facet_grid(. ~ gender) +
  scale_fill_manual(
    values = c(
      "Normal" = "#56B4E9",
      "Elevated" = "#E69F00",
      "Stage 1" = "#D55E00",
      "Stage 2" = "#CC79A7"
    )
  ) +
  labs(
    title = "Counts of SBP categories by age and gender",
    x = "Age group",
    y = "Count",
    fill = "Average SBP"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The complete-case filter means that the distributions describe participants
with age, gender, and at least one systolic blood pressure reading. Faceting and
the ordered colour palette make the age and gender patterns easier to compare.

